---
title: Estimations of Soil Depth with aqp
author: Andrew G. Brown
date: '2020-07-13'
slug: estimations-with-algorithms-for-quantitative-pedology
categories:
  - aqp
  - R
tags:
  - soil taxonomy
  - correlation
---

### Estimation of soil depth in _aqp_ **R** package

Estimation functions for soils try to calculate unknown properties based on existing observations and known relationships. The term pedotransfer function was coined by Johan Bouma as "translating data we have into what we need." Soil surveys, and especially the primary descriptions collected by field soil scientists, offer many opportunities for leveraging this type of information.

In Algorithms for Quantitative Pedology _aqp_, the heuristics are  defined in terms of **R** code. 

There are a variety of functions in _aqp_ that do  "estimation" (several have `estimate` in the name). These functions form a basis for logic underlying many of the high-level functions involving the _SoilProfileCollection_ (SPC) object -- the class for soil data offered by the _aqp_ package. 

Here in this blog post I will talk about one of the most fundamental _aqp_ estimation functions: `estimateSoilDepth`. It uses regular expression pattern matching on horizon designations to estimate depth to a root-limiting layer. If you think creatively, you can do much more by modifying default arguments. Other functions in the same "family" as `estimateSoilDepth` include: `depthOf`, `minDepthOf` and `maxDepthOf`.

Root-limiting layers are relevant to soil classification and interpretations. Horizon designations correlate with the properties that are required for diagnostic contacts, restrictions, etc. so they stand to be a decent surrogates for property functions over depth. 

In the case of identifying contacts, horizon designations are useful and mostly accurate. Though, there are important deviations that are discussed later.

```{r estimateSoilDepth, message=FALSE, warning=FALSE, fig.width=10}
library(aqp)

# load example data from aqp package
data("sp3")

# create a SoilProfileCollection
# depths specifies the profile ID, top and bottom depths in horizon data
depths(sp3) <- id ~ top + bottom

# inspect first 7 profiles from the set
plotSPC(sp3[1:7,])

# aqp::profileApply and aqp::estimateSoilDepth 
# 
# - calculate depth of each profile in sp3
# 
sp3$depth <- profileApply(sp3, estimateSoilDepth)

# aqp::guessHzDesgnName
# 
# helper function to look for regular expression "name" and take first result
# this is historic behavior of aqp::plotSPC and is used by estimateSoilDepth 
# to locate the column containing horizon designations in the horizons slot
# 
guess <- guessHzDesgnName(sp3)
guess

# inspect
horizons(sp3)[[guess]]
```

At the most basic level in an "estimate" you have one or more inputs producing a single, possibly complex, output. Estimates might be made at the horizon, site/profile or group level. 

Let's look at all of the soil depths in the `sp3` collection as a histogram. We make an estimate for each profile, then we summarize those estimates for the "group" `sp3`.

```{r plot-hist-depth, message=FALSE, warning=FALSE}
hist(sp3$depth, 
     main = "Histogram of Soil Depth [sp3]", 
     xlim = c(0,300),
     breaks = 8, freq = TRUE)
```

```{r qtile-density-depth, message=FALSE, warning=FALSE}
res <- quantile(sp3$depth, probs = c(0.05,0.5,0.95))
```

We can summarize distributions of samples and populations ("groups") quantitatively using _quantiles_, as shown in the code above. 

`res` contains the following:

```{r}
res
```

These quantiles are `x` "positions" along an the "probability density function" -- an empirically defined, non-parametric curve fit to the available data. The observed values (`prop`) are the `y` axis.

Approximately 5% of data values are greater than the 95th percentile (0.95 probability level). And likewise for the proportion of values that are less than the 5th percentile. Using the default quantile method, values returned may be interpolated between observed values, depending on the distribution and probability levels requested (`probs` argument).

```{r plot-density-depth, message=FALSE, warning=FALSE, fig.width=10}
# make a plot of soil_depth probability density function (pdf)
plot(density(sp3$depth, na.rm=TRUE, bw=5), 
     main="Density plot of Soil Depth [sp3]")

# add summary statistics
abline(v = res[2], lwd = 2, col = "RED") # median / 50%
abline(v = mean(sp3$depth), lwd = 2, col = "blue") # mean
abline(v = res[-2], lty = 2) # everything other than the median (5%+95%)

# add legend
legend("topright", 
       legend = c("Median","Mean","Q05 + Q95"),
       col = c("RED","BLUE","BLACK"), 
       lty = c(1,2,2), lwd = c(2,2,1))
```

This small dataset of `r length(sp3)` pedons has a unimodal appearance.

The 5th and 95th percentiles generally encompass the moderately deep depth class. The mean is affected by several large values, whereas the median reflects a general concentration of data towards the lower end of the scale.

```{r}
getSoilDepthClass(sp3)
```

Looking back at the original profile plot, we notice most do not have bedrock layers as part of the description though there is wide variation in depth class from moderately to very deep. It would be interesting to have more detailed description about the nature of those contacts.

#### Let's consider Issue #27 from _aqp_.

_aqp_ issue #27 (https://github.com/ncss-tech/aqp/issues/27) is an old issue about densic materials.  I think we should probably close it... since this is a known shortcoming of using horizon designations... but the demo data and code are very useful. Using horizon designations to estimate soil depth has always been prone to the risks of assuming diagnostic criteria _loosely associated_ with those designation are met. Materials below a densic contact pose the greatest risk in terms of contacts as they is most similar to true soil material. 

First, we will replicate the density plot we did above with this the more extensive dataset from issue #27. We will build a `soilDB::fetchNASIS`-like SPC from CSV files that you can download here: [site](csv/20200713/tx_cd_st.csv), [horizon](csv/20200713/tx_cd_hz.csv).

```{r include=FALSE}
library(aqp)
# paths to csv for local build
sit <- read.csv("../../static/csv/20200713/tx_cd_st.csv")
spc <- read.csv("../../static/csv/20200713/tx_cd_hz.csv")
```

```{r eval=FALSE}
library(aqp)

# read in flat CSV files (derived from soilDB::fetchNASIS)
sit <- read.csv("path/to/csv/tx_cd_st.csv")
spc <- read.csv("path/to/csv/tx_cd_hz.csv")
```


```{r}
# promote horizon data to SPC
depths(spc) <- peiid ~ hzdept + hzdepb

# join in site data
site(spc) <- sit

# compute soil depth and depth class for all pedons
# using default regex pattern: 'Cr|R|Cd' from estimateSoilDepth
sdc <- getSoilDepthClass(spc)

# sdc is a data.frame with many useful site level classifications of depth
head(sdc)

# join depth + class data to site
site(spc) <- sdc

# make a plot of depth probability density function (pdf)
plot(density(spc$depth, na.rm = TRUE, bw = 5), 
     main = "Density plot of Soil Depth [Texas Densics]")

# calculate 10th, 50th, 90th percentiles
res <- quantile(spc$depth, probs = c(0.1, 0.5, 0.9))

# add summary statistics
abline(v = res[2], lwd = 2, col = "RED")
abline(v = mean(spc$depth), lwd = 2, lty = 2, col = "blue")
abline(v = res[-2], lty = 2)

# add legend
legend("topright", 
       legend = c("Median","Mean","Q05/Q95"),
       col = c("RED","BLUE","BLACK"), 
       lty = c(1,2,2), lwd = c(2,2,1))
```

#### Densic contacts, densic materials and Cd horizons

The relevant determinations for "densic contact" are the (non-)cementation* (*soon to be renamed to "coherence") class and the evidence of root spacing/limitation. Diagnostic features, and other site-level characteristics that are either observed or inferred, are critically important to soil property estimation _in general_ as well as assessing the "quality" of _particular_ estimates. 

Densic materials and Cd horizons by definition are "physically root-restrictive" -- but they are only _root-limiting layers_ (have a densic contact at their upper bound) if root entry is limited to spaces greater than 10 cm apart. 

In Issue #27 Dylan deals with this in his example code by using the depths in the database for diagnostic "contacts." In my view this is "fine" for a one-off analysis but is of intrinsically lower quality for inference about contacts than the observations "with data." In some cases all we "have" is a decision that the contact occured at a particular depth -- and the most high resolution estimate we have of that is the horizon depths.

At a minimum, the soil scientist writing contact depth on the Form 232 can be used to support a call.

#### What do we do to reconcile the variety of ways soils have been described in space and time?

When making an estimate these questions can be asked for each profile:

1. Do ancillary data "support" the horizon data estimate? 

2. Are there missing data or logic problems precluding use of any data?

3. What is most likely estimate [accounting for 1 and 2]? 

In theory, a "rating" can be made for _each possible_ estimate. The specific types of "estimate support" and logic problems, and their relative weights, are _data-model specific_. But the broad concept of "estimate quality" is generic: stack up a bunch of logical/numeric evaluations on attributes to assign a "quality ranking." 

"Estimates" should be based on one or more site or horizon level properties reduced to a single value per horizon or profile. 

When estimating a site-level value there could be several possible "candidate values" due to many:one attribute:horizon:site relationships. In a single horizon with multiple attributes the aggregate of attributes may favor one value over others, which in turn may affect some aggregation to a site-level value. Where there is "confusion" about which "candidate value" to support the "quality" of the estimate is lower. 

#### Paralithic contacts, paralithic materials, lithic contacts, and the materials below a lithic contact

A common thing I am suspicious of in my area of the world are 25cm thick Cr horizons with no underlying R. This incidentally falls right on the line for the series control section. Was a lithic contact encountered at 25cm that caused them to stop digging? Was it intended to imply that the paralithic materials were thick? Was it a misunderstanding of the difference between describing the whole series control section and a data population standard? 

Digging to 25cm with extreme effort just to call it a 25cm thick Cr defeats the purpose -- so the soil scientist was not  clear on why they were digging / required to describe that far. These observations often are great in terms of soil properties for the solum but are inadequate to assess paralithic materials thickness in the Series Control Section. It is quite likely that in the parent materials in question these pits were not dug to 25cm below the contact. If the Cr is that "diggable," it is not difficult to put it past the threshold for the Series Control Section. If it is not, you should be either describing an R or making a note that the Cr is definitively >25cm thick.

For series criteria, one needs to "decide" whether the soil has a thick or a thin paralithic material "mantle" one way or another. There is nothing magical about "25," other than it is the limit that was chosen. On some level, the decision MUST happen in the field. If adequate ancillary data are recorded in the field, such as: excavation difficulty, cementation, rooting patterns, MOISTURE patterns, then that decision can be assessed to varying degrees in the future.

Deviation between horizons, diagnostics and restrictions is handled variably in Quality Assurance around the country; some of which has probably spurred arbitrary "corrections" of data that were "inconsistent" or missing. In my work I have been allowed to have these properties deviate if needed -- but do not generally feel the need to take this option. 

#### Effects on soil correlation

In my opinion, if you are describing a bunch of Cd horizons whose upper bounds _aren't_ densic contacts/densic restrictions -- you better be recording a bunch of ancillary data to support why you did that! 

Two easy justifications for this in my mind are close root spacing or significant pedogenic development -- which preclude densic materials. Practical reasons why one might "want" a densic restriction despite one or more criteria being met would be to use it to justify Endo- versus Epi aqu- Great Groups. Since up to 25 cm below certain root limiting layers is included in the control section, it can be used to demonstrate "dry materials" within the lower part of the control section. 

This all reflects a much more general "problem" in soil survey: It is very difficult to "prove" the existence of soil types that hinge on recording very specific attributes. This is especially true if the data are not all collected by the same scientist(s) or under the same set of working standards -- but applies just as well because of "Nature." 

Matching up to a very detailed "prototype" pedon doesn't usually pan out when you "fan out all the descriptions for a soil component and squint your eyes." A description containing multiple horizons that could construed as a contact is inherently reflective of a detailed description. 

There is value in having a high-detail in individual descriptions but it requires considering properties (like contacts) "in aggregate" after some sort of risk-weighting logic, as well as other simplifications to make estimates more generic.

#### Extending `estimateSoilDepth`

I would like to be able to leverage more horizon level attributes (such as roots, structure, cementation) for assessing soil depth -- but not as part of `estimateSoilDepth`. 

Switching between name fields for a SPC is easy with `hzdesgnname()` or the relevant `hzname` arguments -- so a single analysis can feasibly use several of these on the same SPC. This extends the ability to use arbitrary character matching for identifying depths to various features. The sky is the limit in terms of the types of data you incorporate into that process. 

```{r}

# calculate shallowest depth of each of the following patterns
depth_to_carbonates <- profileApply(spc, estimateSoilDepth, p = "k")
depth_to_kk  <- profileApply(spc, estimateSoilDepth, p = "kk")
depth_to_gypsum <- profileApply(spc, estimateSoilDepth, p = "y")
depth_to_Cd <- profileApply(spc, estimateSoilDepth, p = "Cd", 
                            no.contact.depth = 225, no.contact.assigned = 250)
depth_to_Cd2 <- profileApply(spc, estimateSoilDepth, p = "Cd[^ky]", 
                            no.contact.depth = 225, no.contact.assigned = 250)
depth_to_Cr <- profileApply(spc, estimateSoilDepth, p = "Cr", 
                            no.contact.depth = 225, no.contact.assigned = 250)

# make a plot showing their depths
plot(main = "Depth to k + y + R|Cr|Cd Horizon Designations",
     density(depth_to_carbonates,  from = 0, bw = 10, na.rm = TRUE), col = 7, lwd = 2)
lines(density(depth_to_kk, from = 0, bw = 10, na.rm= TRUE), col = 3, lwd = 2)
lines(density(depth_to_gypsum, from = 0, bw = 10, na.rm = TRUE), col = 12, lwd = 2)
lines(density(depth_to_Cd, from = 0, bw = 10, na.rm = TRUE), col = 8, lwd = 2)
lines(density(depth_to_Cr, from = 0, bw = 10, na.rm = TRUE), col = 1, lwd = 1, lty = 2)
legend("top", c("k","kk","y","Cd", "Cr"), col = c(7,3,12,8,1), lwd = c(2,2,2,2,1), lty=c(1,1,1,1,2))
```

My preference from here would be to leave `estimateSoilDepth` alone, and generalize `depthOf` further for arbitrary logical expressions like several other functions in `aqp`, only for horizon data explicitly. Pattern matching behavior could be replicated using `grepl` based expressions.

The old dataset I provide in the CSV files that was queried from NASIS before cementation and other criteria were included in the default `fetchNASIS` result. Dylan already did the diagnostic horizon based analysis in his post on the issue, so we will do something different.

Here we inspect for evidence of pedogenesis -- as you cannot meet the criteria for densic materials if you have significant pedogenesis. We can clearly see that all Cd horizons without `k` or `y` suffixes (`depth_to_Cd2`) are deeper than the corresponding estimate that looked for any horizon just containing `Cd`: the default part used in `estimateSoilDepth`. A custom pattern to filter to Cd without pedogenesis or Cr or R would be `"Cd[^ky]|Cr|R"`

```{r}
plot(depth_to_Cd ~ depth_to_Cd2)
```


#### Relevant data elements

The problem of the "bottom of the soil" is a worthy one to tackle. Here are some soil concepts / data elements related to estimating soil "depth" that could possibly be considered in a new function that assesses "quality" of depth estimates (given typical data in a NASIS pedon record).
  
- horizon in lieu texture (bedrock, cemented material -- useful for component data)
- horizon cementation class (ties directly to contact type combined with [inference about] roots)
- root-limiting layer presence/absence (diagnostic contacts) - implies "relatively unaltered" by pedogenesis
- restriction presence/absence (used in interpretations)
- "top of horizon" root location (inconsistent, but occasionally meaningful)
- absence of roots (negative support for >10cm spacing)
- structure (massive suggests large structural units, in general)
- series control section bottom depth (beyond a certain depth there is little value to quibbling about the data)
